---
title: "Caching the Inverse of a Matrix"
author: "Hiten Pragji"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.95in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---
\newpage
\section{A primer on matrix inversion}

Let M be a square matrix, that is, a matrix with the same number of rows and columns. Then, the inverse of M, denoted by M^-1^, is defined by

\begin{align}
  MM^{-1} = M^{-1}M = I
\end{align}

where I is the identity matrix, which is any square matrix all of whose elements in the leading diagonal are 1, and all of whose other elements are zeros. Hence,

\[
\begin{pmatrix}
1 & 0 \\
0 & 1
\end{pmatrix}
\]

is known as a 2x2 identity matrix, and

\[
\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{pmatrix}
\]

is known as the 3x3 identity matrix.

The order in which we multiply M and M^-1^ does not matter, which means that M^-1^, if it exists, is unique.

\section{Assignment}

Matrix inversion is usually a costly computation and there may be some benefit to caching the inverse of a matrix rather than computing it repeatedly (there are also alternatives to matrix inversion that we will not discuss here). Your assignment is to write a pair of functions that cache the inverse of a matrix.

Write the following functions:

1. makeCacheMatrix: This function creates a special "matrix" object that can cache its inverse.
2. cacheSolve: This function computes the inverse of the special "matrix" returned by makeCacheMatrix above. If the inverse has already been calculated (and the matrix has not changed), then cacheSolve should retrieve the inverse from the cache.

Computing the inverse of a square matrix can be done with the solve function in R. For example, if X is a square invertible matrix, then solve(X) returns its inverse.

For this assignment, assume that the matrix supplied is always invertible.

\clearpage

\section{Solution}

\subsection{makeCacheMatrix}

We need to create a function that can create a special object which stores a matrix and caches its inverse. This function performs the following steps:

1. Sets the value of the matrix
  + creates a function called set_matrix
2. Obtains the value of the matrix
  + creates a function called get_matrix
3. Sets the value of the inverse
  + creates a function called set_inverse_value
4. Obtains the value of the inverse
  + creates a function called get_inverse_value

```{r makecachematrix}
makeCacheMatrix <- function(x = matrix()) {
  # Initialise an empty variable to store the cached value of the inverse
  i <- NULL
  # Create a function to set the matrix
  set_matrix <- function(y) {
    x <<- y
    i <<- NULL
  }
  # Create a function to get the matrix
  get_matrix <- function() {
    x
  }
  # Create a function to set the value of the matrix inverse
  set_inverse_value <- function(inverse) {
    i <<- inverse
  }
  # Create a function to get the value of the inverse i
  get_inverse_value <- function() {
    i
  }
  # Combine all the function elements into a list
  special <- list(
    set_matrix = set_matrix,
    get_matrix = get_matrix,
    set_inverse_value = set_inverse_value,
    get_inverse_value = get_inverse_value
  )
  # Return the list
  return(special)
}
```

\subsection{cacheSolve}

We need to create a function that computes the inverse of the special matrix returned by makeCacheMatrix and, if the inverse has already been calculated (and the matrix has not changed), then this function should retrieve the inverse from the cache. The steps are as follows:

1. Retrieve the inverse i from makeCacheMatrix:
  + in the event that i has already been calculated and the matrix has not changed
  + in the event that i has not been calculated
2. Retrieve the matrix x
3. Compute the inverse

```{r cachesolve}
cacheSolve <- function(x, ...) {
  # Check if the inverse has already been computed and stored in cache
  i <- x$get_inverse_value()
  # If i has already been calculated, retrieve the inverse
  if (!is.null(i)) {
    message("Retrieving the cached inverse.")
    return(i)
  }
  # If the inverse is null or the cached inverse was invalidated, get a matrix
  m <- x$get_matrix()
  # Calculate its inverse
  i <- solve(m, ...)
  # Store the computed inverse in cache
  x$set_inverse_value(i)
  # Return the computed inverse
  return(i)
}
```

\subsection{Applying the functions}

```{r calculate_inverse}
# Use the makeCacheMatrix function to create a special matrix object
M <- makeCacheMatrix(x = matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2))
# Print the matrix M
print(M$get_matrix())
# Compute the matrix inverse using the cacheSolve function
N <- cacheSolve(M)
# Print the inverse N
print(N)
```



















